Massively Parallel Monte Carlo Methods for Linear and Nonlinear Problems

### 2012-10-02 doc/prelim/document/chapters/research_proposal/research_proposal.tex
added outline of slides for prelim

\chapter{Research Proposal}
\label{ch:research_proposal}
To develop the parallel MCSA and FANM methods, a research plan is
proposed to drive their development and quantify their merits and
feasibility as compared to more conventional methods for linear and
nonlinear problems. We seek to rigorously verify both of these methods
and there implementation and will describe the means by which that
will be accomplished. Furthermore, the numerical experiments performed
will be constructed within a software framework that will permit an
agile development strategy. In this chapter we describe that
experimental framework and the progress achieved to date by its
use. In addition, we will a formulate a plan for verifying the new
stochastic methods and prepare a set of numerical experiments that
will determine some of their properties. Finally, we select a
large-scale challenge problem designed to potentially demonstrate the
effectiveness of our new methods in the context of real-world,
production level applications for nuclear reactor analysis.

@@

In our experimental framework, we will formulate all of our problems
using the weak form of the differential equations as in
Eq~(\ref{eq:fe_weak_form}). Doing so allows us to leverage modern
finite element assembly engine software with embedded automatic
differentiation technology. Using the Panzer library in the Trilinos
scientific computing framework
\citep{notz_graph-based_2010,heroux_overview_2005}, function
evaluation routines in the form of Eq~(\ref{eq:fe_weak_form}) are
generated to represent the finite element problem and boundary
conditions. Per Bartlett's automatic differentiation work and an
associated parallel linear algebra framework, the equation sets for
all physics implemented are assembled into the nonlinear system. By
leveraging the general linear algebra framework, we are able to
abstract away those concepts in the stochastic solvers implementation
and instead focus on an algorithm-oriented approach for the solver
implementation \citep{musser_algorithm-oriented_1994}.

@@

\section{Stochastic Methods Verification}
\label{sec:method_verification}
As we perform our numerical experiments, the results from the new
tools that we are using must be verified against benchmark
results. These benchmark tests should allow use to verify that the new
stochastic solvers that we develop generate the expected numerical
results in parallel for linear and nonlinear problems that have a
known, well-documented solution. In addition, as we explore parallel
strategies for the Monte Carlo methods and introduce bias or
non-reproducible results into our solution estimators, these
verification problems will allow us to determine if that cost is
acceptable. To verify the linear solvers, we will choose the above
2-dimensional heat equation problem in Eq~(\ref{eq:heat_equation}) on
a rectilinear grid for which analytical solutions are available. As a
complement to comparison with the analytical solutions, we may also
check those results against production implementations of Krylov
solvers. For the nonlinear solvers, we employ a new set of equations
with readily available benchmark solutions.

>>>>

\chapter{Research Proposal}
\label{ch:research_proposal}
To develop the parallel MCSA and FANM methods, a research plan is
proposed to drive their development and quantify their merits and
feasibility as compared to more conventional methods for linear and
nonlinear problems. We seek to rigorously verify both of these methods
and there implementation and will describe the means by which that
will be accomplished. Furthermore, the numerical experiments performed
will be constructed within a software framework that will permit an
agile development strategy. In this chapter we describe that
experimental framework and the progress achieved to date by its
use. In addition, we will a formulate a plan for verifying the new
Monte Carlo methods and prepare a set of numerical experiments that
will determine some of their properties. Finally, we select a
large-scale challenge problem designed to potentially demonstrate the
effectiveness of our new methods in the context of real-world,
production level applications for nuclear reactor analysis.

@@

In our experimental framework, we will formulate all of our problems
using the weak form of the differential equations as in
Eq~(\ref{eq:fe_weak_form}). Doing so allows us to leverage modern
finite element assembly engine software with embedded automatic
differentiation technology. Using the Panzer library in the Trilinos
scientific computing framework
\citep{notz_graph-based_2010,heroux_overview_2005}, function
evaluation routines in the form of Eq~(\ref{eq:fe_weak_form}) are
generated to represent the finite element problem and boundary
conditions. Per Bartlett's automatic differentiation work and an
associated parallel linear algebra framework, the equation sets for
all physics implemented are assembled into the nonlinear system. By
leveraging the general linear algebra framework, we are able to
abstract away those concepts in the Monte Carlo solvers implementation
and instead focus on an algorithm-oriented approach for the solver
implementation \citep{musser_algorithm-oriented_1994}.

@@

\section{Monte Carlo Methods Verification}
\label{sec:method_verification}
As we perform our numerical experiments, the results from the new
tools that we are using must be verified against benchmark
results. These benchmark tests should allow use to verify that the new
stochastic solvers that we develop generate the expected numerical
results in parallel for linear and nonlinear problems that have a
known, well-documented solution. In addition, as we explore parallel
strategies for the Monte Carlo methods and introduce bias or
non-reproducible results into our solution estimators, these
verification problems will allow us to determine if that cost is
acceptable. To verify the linear solvers, we will choose the above
2-dimensional heat equation problem in Eq~(\ref{eq:heat_equation}) on
a rectilinear grid for which analytical solutions are available. As a
complement to comparison with the analytical solutions, we may also
check those results against production implementations of Krylov
solvers. For the nonlinear solvers, we employ a new set of equations
with readily available benchmark solutions.

### 2012-10-02 doc/prelim/presentation/prelim_presentation.tex
added outline of slides for prelim

\begin{frame}{My Research and Development Activities}

  \begin{itemize}
  \item Stochastic computational linear algebra.
  \item Multiphysics methods for reactor physics.
  \item High performance computing.
  \item Software Engineering.
  \end{itemize}

@@

%%---------------------------------------------------------------------------%%
\begin{frame}{Data Transfer Kit (DTK)}

  DTK is a software component for generating and applying parallel
  topology map operators based on multiphysics parameters. Some
  requirements:

  \begin{itemize}
  \item A component based design not limited to specific mesh or
    field data structures
  \item Compatible with both mesh-based and mesh-free physics codes
  \item Based on load-balanced rendezvous algorithms of a desirable
    time complexity
  \item Operate on $O(1,000,000,000)$ element meshes
  \item Operate on $O(100,000)$ cores
  \item Integrate readily with current CASL physics components
  \item Open-source implementation (DTK has a BSD 3-clause license)
  \end{itemize}

@@

%%---------------------------------------------------------------------------%%
\begin{frame}{Stochastic Linear and Nonlinear Solvers}

  Ph.D. Research: Massively Parallel Stochastic Methods for Linear and
  Nonlinear Problems.

  \begin{itemize}
  \item Moving to exascale facilities requires methods tolerant to
    both soft and hard errors
  \item Solver scalability of primary concern
  \item Difficult to form the Jacobian analytically for block operator
    problems, Jacobian-free methods are too coarse or do not perform
    well for poorly scaled problems
  \item Subspace (e.g. Krylov) methods are memory intensive
  \item Stochastic solvers are a novel research area that may bear
    fruit
  \item We can leverage bleeding-edge Monte Carlo methods in reactor
    physics for parallelism
  \end{itemize}

@@

%%---------------------------------------------------------------------------%%
\begin{frame}{MCSA and FANM}

  \textbf{Monte Carlo Synthetic Acceleration (MCSA)}: A stochastic
  solution method for sparse linear problems.

  \begin{subequations}
    \label{eq:MCSA-iteration}
    \begin{gather}
      \vx^{l+1/2} = (\vI - \vA)\vx^l + \vb\:,\\ \vr^{l+1/2} = \vb -
      \vA\vx^{l+1/2}\:,
      \label{eq:MCSA-residual}\\     
      \hat{\vA}\delta\vx^{l+1/2} = \vr^{l+1/2}\:,
      \label{eq:MCSA-residual_solve}\\ 
      \vx^{l+1}=\vx^{l+1/2}+\delta\vx^{l+1/2}\:.
    \end{gather}
  \end{subequations}

  \textbf{Forward-Automated Newton-MCSA (FANM)}: A stochastic solution
  method for nonlinear problems

  \begin{itemize}
  \item Form the Jacobian explicitly with forward automatic
    differentiation. (No nasty derivatives, no Jacobian
    approximations, compressed row storage cheaper than the subspace
    over many iterations)
  \item Use MCSA as the linear solver. (All the benefits of the
    stochastic solver)
  \end{itemize}

@@

%%---------------------------------------------------------------------------%%
\begin{frame}{General MCSA Preliminary Results}

  Transient Poisson equation

  \begin{figure}[htpb!]
    \begin{center}
      \includegraphics[width=4in]{basic_results.png}
    \end{center}
  \end{figure}

  \begin{itemize}
  \item MCSA implemented in C++ using Epetra linear algebra
    framework. General linear algebra implemenation possible with
    MCSA!
  \item Competitive and GMRES and Conjugate Gradient solvers.
  \end{itemize}

  An initial FANM implementation is close and will be based on the
  Epetra solver. Parallel MCSA will be later in the fall per
  development of the Chimera code base.

>>>>

\begin{frame}{Introduction}

@@

%%---------------------------------------------------------------------------%%
\begin{frame}{Physics-Based Motivation}

@@

%%---------------------------------------------------------------------------%%
\begin{frame}{Hardware-Based Motivation}

@@

%%---------------------------------------------------------------------------%%
\begin{frame}{Research Outline}

@@

%%---------------------------------------------------------------------------%%
\begin{frame}{Conventional Solution Methods for Linear Problems}

%%---------------------------------------------------------------------------%%
\begin{frame}{Linear Problem Preliminaries}

%%---------------------------------------------------------------------------%%
\begin{frame}{Stationary Methods}

%%---------------------------------------------------------------------------%%
\begin{frame}{Projection Methods}

%%---------------------------------------------------------------------------%%
\begin{frame}{Parallel Projection Methods}

%%---------------------------------------------------------------------------%%
\begin{frame}{Monte Carlo Solution Methods for Linear Problems}

%%---------------------------------------------------------------------------%%
\begin{frame}{Monte Carlo Linear Solver Preliminaries}

%%---------------------------------------------------------------------------%%
\begin{frame}{Direct Method}

%%---------------------------------------------------------------------------%%
\begin{frame}{Adjoint Method}

%%---------------------------------------------------------------------------%%
\begin{frame}{Sequential Monte Carlo}

%%---------------------------------------------------------------------------%%
\begin{frame}{Monte Carlo Synthetic-Acceleration}

%%---------------------------------------------------------------------------%%
\begin{frame}{Parallelization of Stochastic Methods}

%%---------------------------------------------------------------------------%%
\begin{frame}{Monte Carlo Solution Methods for Nonlinear Problems}

%%---------------------------------------------------------------------------%%
\begin{frame}{Monte Carlo Nonlinear Solver Preliminaries}

%%---------------------------------------------------------------------------%%
\begin{frame}{Inexact Newton Methods}

%%---------------------------------------------------------------------------%%
\begin{frame}{The FANM Method}

%%---------------------------------------------------------------------------%%
\begin{frame}{Research Proposal}

%%---------------------------------------------------------------------------%%
\begin{frame}{Experimental Framework}

%%---------------------------------------------------------------------------%%
\begin{frame}{Progress to Date}

%%---------------------------------------------------------------------------%%
\begin{frame}{Monte Carlo Methods Verification}

%%---------------------------------------------------------------------------%%
\begin{frame}{Proposed Numerical Experiments}

%%---------------------------------------------------------------------------%%
\begin{frame}{Proposed Challenge Problem}

%%---------------------------------------------------------------------------%%
\begin{frame}{Conclusion}
